---
title: "Getting started with warwickplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with warwickplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
ragg_png <- function(..., res = 300) {
  ragg::agg_png(..., res = res, units = "in")
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png"
)
```

```{r setup}
library(warwickplots)
```

warwickplots is an R package with colour palettes and a [ggplot2](https://ggplot2.tidyverse.org) theme that are consistent with The University of Warwick's [branding](https://warwick.ac.uk/about/brand/brand-guidelines/),
especially its [colours](https://warwick.ac.uk/about/brand/brand-guidelines/colours/) and [typography](https://warwick.ac.uk/about/brand/brand-guidelines/typography/).

The palettes are built using the [palettes](https://mccarthy-m-g.github.io/palettes/index.html) package.
The ggplot2 theme is heavily inspired by the work of [Cara Thompson](https://www.cararthompson.com/), particularly this talk on [applying a unifying aesthetic to your plots](https://www.cararthompson.com/talks/nhsr2022-ggplot-themes/).

## The palettes

The package offers six colour palettes: one discrete, two divergent and three sequential:

```{r palettes}
library(warwickplots)
plot(warwick_palettes)
```

`warwick_palettes` is essentially a list, and so individual palettes can be accessed by name:^[In the RStudio IDE and in the [online vignette](https://warwick-stats-resources.github.io/warwickplots/articles/warwickplots.html), the print method for palettes shows each bullet point in the relevant colour, but in the vignette built for R, all the bullet points are the same (blue) colour.]

```{r}
warwick_palettes$primary
```
Each palette is essentially a vector, so can be subsetted in the usual way to create new palettes:

```{r}
warwick_palettes$primary[c(1, 3, 5)]
```
The palettes package provides a `palettes::pal_ramp()` function for interpolating between colours. For example, consider the `warwick_palettes$aubergine` palette:

```{r}
plot(warwick_palettes$aubergine)
```

To generate a new palette of five colour values between the darkest and mid-point of the full palette, and reverse it (then visualise it), we can do the following:

```{r}
ends <- warwick_palettes$aubergine[c(1, 4)] # get new end-points
pal_ramp(ends, n = 5, direction = -1) |> plot()
```

For more on methods for working with palettes, see the [palettes](https://mccarthy-m-g.github.io/palettes/articles/palettes.html) vignette in the palettes package.

## Using palettes in ggplot2

The palettes package provides colour and fill scales for using its palettes in ggplot2. These are:

- `scale_colour_palette_d()`
- `scale_colour_palette_c()`
- `scale_colour_palette_c()`
- `scale_fill_palette_d()`
- `scale_fill_palette_c()`
- `scale_fill_palette_c()`

The `_d` variants are for discrete scales, the `_c` variants are for continuous scales and the `_b` variants are for binned scales.


See the [using warwickplots with ggplot2](https://warwick-stats-resources.github.io/warwickplots/articles/ggplot2.html) vignettes for more information and examples and also the [using palettes with ggplot2](https://mccarthy-m-g.github.io/palettes/articles/ggplot2.html) vignette in the palettes package. The palettes package also has vignettes demonstrating how to use palettes with the [gt](https://gt.rstudio.com) package and the [biscale](https://chris-prener.github.io/biscale/) package.

## Typography and setting up custom fonts

The University of Warwick's [typography brand guidance](https://warwick.ac.uk/about/brand/brand-guidelines/typography/) is to use the font Lato for all online text and the font Avenir Next for all print material. `theme_warwick()` has a `use` argument, that can be one of `"online"` or `"print"` (defaults to `"online"`), which will ensure the appropriate font for the use, *if* your system is set up for it, i.e. you have the fonts installed and the packages to render them in a plot. This section contains guidance on making sure your system is suitably set-up.

### Setting up ragg

You need to have the [ragg](https://ragg.r-lib.org), [systemfonts](https://github.com/r-lib/systemfonts), and [textshaping]() packages installed. Installing ragg should install the other two:

```{r eval = FALSE}
install.packages('ragg')
```

You then need to ensure that whenever you output a plot, you do so using the **AGG** graphics device. 

- In RStudio >= 1.4, go to Tools -> Global Options -> General -> Graphics and set the Backend to AGG.

- In Quarto, set the following YAML:^[This is already set as the default in [warwickpres](https://github.com/Warwick-Stats-Resources/warwickpres)]

```
  knitr:
    opts_chunk: 
      dev: "ragg_png"
```

- In RMarkdown, use `knitr::opts_chunk$set(dev = "ragg_png")` at the top of a script.

### Checking the fonts are installed

To use either of the fonts, you first need to check if they are installed on your system:

```{r}
systemfonts::match_fonts(c("Lato", "Avenir Next"))
```

If the paths in the tibble show Lato and Avenir Next, then they are installed on your system. If they show another font, i.e. your system default, then they are not. For example, in the environment in which this vignette is rendered, there is access to Lato, but not to Avenir Next. If I tried to set the font family for a plot to Avenir Next, it would display the default DejaVuSans instead.

If you do not have Lato installed, you can install it, either directly from [Google Fonts](https://fonts.google.com/specimen/Lato?query=Lato) or using the [sysfonts](https://github.com/yixuan/sysfonts) package:

```{r eval = FALSE}
sysfonts::font_add_google("Lato")
```

For further excellent, detailed guidance on setting up and working with custom fonts in R, see the [getting fonts to work in R](https://www.cararthompson.com/posts/2024-01-12-using-fonts-in-r-for-dataviz/2024-01-12_getting-fonts-to-work#set-your-graphics-device-to-agg) post by Cara Thompson and the [setting up and debugging custom fonts](https://yjunechoe.github.io/posts/2021-06-24-setting-up-and-debugging-custom-fonts/) post by June Choe.
